One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-26 paper/discussion.tex
minor updates
second draft

% According to previous studies \citep{adams89}, our disc models
% should not support $m=1$ linear spiral instabilities. This is because
% our discs are not sufficiently massive ($M_d\lesssim 0.1M_*$) and, more
% importantly, we have surpressed the motion of the central star. The 
% latter can induce $m=1$ instabilities in massive discs  \citep{shu90}.

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes generally have negative
angular momentum. If the mode loses (positive) angular momentum
to the background, then we can expect instability. We demonstrate
this below.

@@

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align*}
  T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
\end{align*}
The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} is negligible. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R,
\end{align}
which gives
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
\end{align}
% \begin{align}
%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{\sbar}\right)^*\right]
% \end{align}
% where we replaced the radial Lagrangian displacement by the Eulerian
% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
% The linearised momentum equations give
% \begin{align}
%   D\delta v_{Rm} =& 
%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
%   &- \frac{2\ii
%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
%     \delta\Phi_m\right),
% \end{align}
% where $D\equiv \kappa^2 - \sbar^2$. 
% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
% real, and using the local solution to the Poisson equation
% \begin{align}
%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
% \end{align}
% \citep{shu91}. Then the radial velocity perturbation becomes
% \begin{align}
%   \delta v_{Rm} =
%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
%       m\Omega}{R}\right),
% \end{align}
% so that 
% \begin{align}
%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
%       m\Omega}{R\sbar^*}\right). 
% \end{align}
% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
% taking the imaginary part of the above expression, the torque
% density is 
% \begin{align}\label{tbg_explicit}
%   T_\mathrm{BG} &=
%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\right].
% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
% \end{align}
% We now use the local dispersion relation,
% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
% k^2c_s^2$ in the above expression. This gives
% \begin{align}\label{tbg_simple}
%   T_\mathrm{BG} =
%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
% \end{align}  
This torque density is negative for trailing waves in discs with
a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
    \xi_\phi|^2}. 
%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
2\ii \xi_R$ for the modes of interest. Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
For perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
our simulations, we have 
\begin{align}\label{theoretical_rate}
  \gamma \sim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
consistent with simulation results.

%Of course, we assumed a neutral mode to obtain expressions
%for $\jlin$ and $T_\mathrm{BG}$ in the first place, 
Note that by using local theory we have implicitly
assumed a neutral mode because the dispersion relation,
Eq. \ref{dispersion}, states that non-axisymmetric modes are
stable for real $k$. Eq. \ref{theoretical_rate} should be interpreted
as an order-of-magnitude estimate of the perturbative effect of a
temperature gradient on local, low-frequency $m=1$ modes.

@@

\subsection{The role of self-gravity and disc structure} 
In the above discussion, we only considered a self-gravitating
perturbation to obtain an appropriate wavenumber $k$ needed to
evaluate Eq. \ref{theoretical_rate}. 
The background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity explicitly, so self-gravity is not directly
responsible for instability. However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

>>>>

The dominant disturbance in our simulations is an one-armed, $m=1$
trailing spiral, and its growth is associated with the imposed temperature
gradient. As discussed in \S\ref{global_cons}, in a locally isothermal disc this imposed
temperature gradient generally results in a torque being exerted on
linear perturbations. An instability is therefore possible if the sign
of this torque is the same as that of the angular momentum associated
with the linear perturbation.

%associated angular mometum density is the same as that
%of this
% According to previous studies \citep{adams89}, our disc models
% should not support $m=1$ linear spiral instabilities. This is because
% our discs are not sufficiently massive ($M_d\lesssim 0.1M_*$) and, more
% importantly, we have surpressed the motion of the central star. The 
% latter can induce $m=1$ instabilities in massive discs  \citep{shu90}.

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes generally have negative
angular momentum. If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We demonstrate
below how this is possible through a forced temperature
gradient.

@@

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align*}
  T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
\end{align*}
The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R,
\end{align}
which gives
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
\end{align}
% \begin{align}
%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{\sbar}\right)^*\right]
% \end{align}
% where we replaced the radial Lagrangian displacement by the Eulerian
% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
% The linearised momentum equations give
% \begin{align}
%   D\delta v_{Rm} =& 
%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
%   &- \frac{2\ii
%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
%     \delta\Phi_m\right),
% \end{align}
% where $D\equiv \kappa^2 - \sbar^2$. 
% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
% real, and using the local solution to the Poisson equation
% \begin{align}
%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
% \end{align}
% \citep{shu91}. Then the radial velocity perturbation becomes
% \begin{align}
%   \delta v_{Rm} =
%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
%       m\Omega}{R}\right),
% \end{align}
% so that 
% \begin{align}
%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
%       m\Omega}{R\sbar^*}\right). 
% \end{align}
% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
% taking the imaginary part of the above expression, the torque
% density is 
% \begin{align}\label{tbg_explicit}
%   T_\mathrm{BG} &=
%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\right].
% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
% \end{align}
% We now use the local dispersion relation,
% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
% k^2c_s^2$ in the above expression. This gives
% \begin{align}\label{tbg_simple}
%   T_\mathrm{BG} =
%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
% \end{align}  
This torque density is negative for trailing waves in discs with
a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate 
the growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
    \xi_\phi|^2}. 
%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
2\ii \xi_R$ for the modes of interest. Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
For perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
our simulations, we have 
\begin{align}\label{theoretical_rate}
  \gamma \sim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
consistent with simulation results. Specifically,
Eq. \ref{theoretical_rate} indicates the growth rate should increase
linearly with $q$ and $h$, as seen in the simulations
(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).

%Note that changing $q$ affects the magntiude of the background
%torque through its explicit depdence on the temperature gradient. By
%definition, the global temperature gradient does not affect modes in
%local theory. 
%however, changing $h$ may also modify the mode properties (since
%temperature appears in local theory) in addition
%to the magnitude of the temperature gradient. Thus, we expect the depedence of
%the background torque on $q$ to be better described by
%Eq. \{theoretical_rate}

%by definition, q does not affect local neutral modes
%Of course, we assumed a neutral mode to obtain expressions
%for $\jlin$ and $T_\mathrm{BG}$ in the first place, 
%Recall that in local theory non-axisymmetric modes are stable for real
%$k$  
%Eq. \ref{theoretical_rate} should be interpreted
%as an order-of-magnitude estimate of the perturbative effect of a
%temperature gradient on local, low-frequency $m=1$ modes.

@@

\subsection{The role of self-gravity and disc structure} 
It is important to note that in the above discussion, we only considered a self-gravitating
perturbation to obtain an appropriate wavenumber $k$ needed to
evaluate Eq. \ref{theoretical_rate}. 
The background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity explicitly, so self-gravity is not directly
responsible for instability.

However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.
